---
title: 'Coursera Reproducible Research Peer Assessment #1'
author: "Bill Cary"
date: "Sunday, December 7, 2014"
output: html_document
---

#### Set defaults
Set knitr to echo code by default
```{r setoptions, echo = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set echo=TRUE by default
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(knitr)
opts_chunk$set(echo = TRUE)

```

#### Import the R libraries required for the analysis
```{r prep}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Import required libraries
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
suppressMessages(library(ggplot2))  # General plotting functions
suppressMessages(library(dplyr)) # Data manipulation
suppressMessages(library(reshape2))
suppressMessages(library(DescTools))
suppressMessages(library(lubridate))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set paths for files and directory structure
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
path_data <- '../data/activity.csv'

```

#### Loading and preprocessing the data
Read the dataset into R and summarize the data
```{r loaddata}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Read the data into R
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
data <- read.csv(path_data)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Briefly summarize the dataset
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Desc(data)
```

#### Miscellaneous Support Information
Print out system and session information to support future readers in
replicating the environment used to produce the output, and in troubleshooting
any potential issues that arise when running the code.
```{r info}
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Print System and Session for reproducibility & troubleshooting.
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
print(sessionInfo())
print(Sys.info())
```