---
title: 'Coursera Reproducible Research Peer Assessment #1'
author: "Bill Cary"
date: "Sunday, December 7, 2014"
output: html_document
---

#### Set defaults
Set knitr to echo code by default
```{r setoptions, echo = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set echo=TRUE by default
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(knitr)
opts_chunk$set(echo = TRUE)

```

#### Import the R libraries required for the analysis
```{r prep}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Import required libraries
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
suppressMessages(library(ggplot2))     # General plotting functions
suppressMessages(library(plyr))        # Data manipulation
suppressMessages(library(dplyr))        # Data manipulation
suppressMessages(library(lubridate))   # Handling date values
suppressMessages(library(Amelia))      # Imputation of missing values

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set paths for files and directory structure
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
path_data <- '../data/activity.csv'

```

#### Loading and preprocessing the data
Read the dataset into R and summarize the data.
```{r loaddata}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Read the data into R
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
data <- read.csv(path_data)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Briefly summarize the dataset
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
summary(data)
str(data)
str(data$steps)
str(data$date)
str(data$interval)
```

#### Transform the Data
The data['date'] field is stored as a factor and needs to be converted to a date.
In addition, the data['steps'] field should be copied over and NA values should be
replaced with real values for use later in the exercise.
```{r transform}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a new feature that stores the data$date field as an actual data value.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
data$date.trans <- ymd(data$date)
str(data$date.trans)

str(data)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Impute the missing values in data$steps
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
data$steps.trans <- mean(data$steps, na.rm = TRUE)
```

#### What is mean total number of steps taken per day?
First, we plot a histogram illustrating the distribution of steps taken per
day.  To do, this we first summarize the data to determine the total number
of steps taken each day.  (Because the raw dataset reports the data in five
minute intervals.)
```{r histogram}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Summarize data to determine the total number of steps taken each day.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
sumdata <- ddply(data, c('date.trans'), summarise,
               dailysum = sum(steps))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a histogram plot
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
qplot(sumdata$dailysum, binwidth = 1350)
```


Next, we calculate the mean and median total steps per day, ignoring NA values.
```{r meanmedian}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Use data summarized above to determine daily mean and median steps
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
meansteps <- mean(sumdata$dailysum, na.rm = TRUE)
mediansteps <- median(sumdata$dailysum, na.rm = TRUE)
```

The mean number of steps taken per day is `r meansteps` and the median number
of steps taken per day is `r mediansteps`.

#### What is the average daily activity pattern?
Next, we will create a time series plot of the average number of steps across
all days (y-axis) by five-minute interval (x-axis).  To create this plot, we
first use plyr to calculate the average number of steps for each interval prior
to creating the time series plot.
```{r dailyactivity}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Summarize data to determine the total number of steps taken each day.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
dailyavg <- ddply(data, c('interval'), summarise,
               dailymean = mean(steps, na.rm = TRUE))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create the plot
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
ggplot(dailyavg, aes(x=interval, y=dailymean)) + 
        geom_point(shape=1)      # Use hollow circles
```

Now that we have averaged the number of steps by interval, it is simple to
determine the specific interval that has, on average, the highest number of
steps.
```{r maxstepsinterval}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Use which.max() to determine the row with the maximum number of steps
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
dailyavg[which.max(dailyavg[,2]),]
```

As can be seen above, the interval with the maximum average number of steps
is interval number `r dailyavg[which.max(dailyavg[,2]),][,1]`.

####Imputing missing values
First, we determine the number of missing values in the steps field as follows:
```{r nummissingvalues}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Summarize the data$steps field to see the number of NA values
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
summary(data$steps)
```
As can be seen above, the steps field contains `r sum(is.na(data$steps))` NA
values.  Next, we will create a new dataset that is a copy of the old dataset,
but with NA values replaced by the average for the respective intervals.
```{r replaceNAs}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Group by interval number, and calculate the interval means
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
interval_means <- summarise(grouped <- group_by(data, interval),
                            intmean = mean(steps, na.rm = TRUE))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Copy the original dataframe, then merge the new interval means dataframe
# with the copy
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
newdata <- data
merged <- merge(newdata, interval_means, "interval")

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Add a new column containing the steps value, with NA's replaced by the
# interval mean
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
merged <- mutate(merged, newsteps = ifelse(is.na(steps), intmean, steps))
```
Following these operations, the new column - newsteps - contains no NA values.
`r summary(merged$newsteps)`

Next, using the new data, we plot a histogram illustrating the distribution of
steps taken per day.  To do, this we first summarize the data to determine the
total number of steps taken each day.  (Because the raw dataset reports the
data in fiveminute intervals.)
```{r newhistogram}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Summarize data to determine the total number of steps taken each day.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
sumdata <- ddply(merged, c('date.trans'), summarise,
               dailysum = sum(newsteps))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a histogram plot
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
qplot(sumdata$dailysum, binwidth = 1350)
```


Next, we calculate the mean and median total steps per day, ignoring NA values.
```{r newmeanmedian}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Use data summarized above to determine daily mean and median steps
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
meansteps <- mean(sumdata$dailysum, na.rm = TRUE)
mediansteps <- median(sumdata$dailysum, na.rm = TRUE)
```

The mean number of steps taken per day is `r meansteps` and the median number
of steps taken per day is `r mediansteps`.

#### Miscellaneous Support Information
Print out system and session information to support future readers in
replicating the environment used to produce the output, and in troubleshooting
any potential issues that arise when running the code.
```{r info}
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Print System and Session for reproducibility & troubleshooting.
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
print(sessionInfo())
# print(Sys.info())   omit Sys.info() for privacy reasons
```